Simulating High-Dimensional Multivariate Data

### 2021-02-25 030-bigsimr-package.Rmd
Revise 030, cut rslurm.

@@

The following description is directed towards towards `R` users via the `bigsimr` `R` package. This package provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` at PROVIDE URL that interfaces with the Julia `Bigsimr` package. The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)). When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize and automate workflows.

The subsections below describe the basic use of `bigsimr`, by stepping through a low-dimensional (2D) simulation workflow for the readily-available data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983]. This workflow proceeds from data, to estimation, simulation configuration, random vector generation, and result visualization. For this low-dimensional setting, we operate using a single central processing unit (CPU), as the overhead in forking the tasks to multiple cores outweighs the computational gains. Then we transition to advanced use where we briefly describe some of the high-performance features and syntax. The section concludes with a short description of how to use `bigsimr` on a computer clusters through slurm scheduling via the `rslurm` package [@R-rslurm].

## Bivariate example of basic use

We demonstrate the basic use and syntax of `bigsimr` through an example workflow applied to the New York air quality data set (`airquality`) included in the R `datasets` package. First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

@@

*Specifying marginal distributions*.
Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed. We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

*Specifying correlation*.
As mentioned, the user must decide how to describe correlation, based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support approximate Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel. (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

@@

*Checking target correlation matrix admissibility*.
First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)). `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

@@

*Simulating random vectors*.
Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

@@

## Advanced use {#advanced-use}

*Using `bigsimr` on a computing cluster via `rslurm`*.
Although `bigsimr` runs quickly, for large $d$ users may wish to run jobs on a shared computing server. The R package `rslurm` enables parallel `rvec` calls. This example assumes that `bigsimr` is installed on a system with a slurm scheduler. A single run of `rvec` using `rslurm` can be coded as:

```{r ch030-rvecRslurm-with-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_single_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
sjob <- slurm_call(bs$rvec, jobname = 'rvec',
                   list(1e6,
                        rho,
                        margins),
                   submit = TRUE)
```

```{r ch030-rvecRslurm, echo=TRUE, eval=FALSE}
library(rslurm)
sjob <- slurm_call(bs$rvec, jobname = 'rvec',
                   list(1e6, 
                        rho, 
                        margins),
                   submit = TRUE)
```

<!-- 
Once your jobs complete, `rslurm` conveniently stores simulation results in a standardized format. Retriving the data is accomplished by
-->

```{r ch030-rvecRslurmRes, echo=FALSE, eval=FALSE}
res <- readRDS(file.path('_rslurm_rvec', list.files('_rslurm_rvec', 'results')))
head(res, 3)
```

The power of combining `bigsimr` and `rslurm` can be illustrated by simulating many correlation structures. The `rslurm::slumr_map` syntax mirrors the `base::lapply` and `purrr::map` functions.

```{r ch030-rvecRslurmMap-comments, echo=FALSE, eval=FALSE}
library(rslurm)
## a single call of rvec
## had to edit rslurm/templates/submit_sh.txt
## source ~/.bashrc
## conda activate bigsimr
## conda env config vars list
set.seed(06202020)
simReps <- 100
rhoList <- replicate(n = simReps, bs::cor_randPSD(length(margins)),
                     simplify=FALSE)
sjob <- slurm_map(x = rhoList,
                  f = bs$rvec,
                  jobname = 'rvecMap',
                  1e6,
                  margins,
                  nodes = 4,
                  cpus_per_node = 4,
                  submit = TRUE)
```

```{r ch030-rvecRslurmMap, echo=TRUE, eval=FALSE}
library(rslurm)
simReps <- 100
rhoList <- replicate(n = simReps, bs::cor_randPSD(length(margins)),
                     simplify=FALSE)
sjob <- slurm_map(x = rhoList,
                  f = bs$rvec,
                  jobname = 'rvecMap',
                  1e6,
                  margins,
                  nodes = 4,
                  cpus_per_node = 4,
                  cores = 4,
                  submit = TRUE)
```

On a cluster carrying 24 nodes with 48 threads, these 100 jobs complete in about a minute.

<!-- 
Let's evaluate the quality of simulation performance.
-->

```{r ch030-rvecRslurmMapRes, echo=FALSE, eval=FALSE}
list.files('_rslurm_rvecMap', 'results')
res <- rslurm::get_slurm_out(sjob, outtype = 'raw')
rhoHats <- lapply(res, bigsimr::fastCor, method = 'spearman')
resData <- NULL
simreps <- length(rhoHats)
for (i in 1:simreps){
    tmpRhoHat <- rhoHats[[i]]
    ## average together
    margins
    avgRhoHat <- tmpRhoHat[lower.tri(tmpRhoHat)]
    tmpPair <- paste0(
      "pair", apply(combn(x = length(margins), 2), 2, paste, collapse = "_")
    )
    tmpResData <- data.frame(
      simNum = i,
      rho    = rhoList[[i]][1,2], 
      rhoHat = tmpRhoHat[lower.tri(tmpRhoHat)], 
      pair   = tmpPair
    )
    resData <- rbind(resData, tmpResData)
}
## "wrangle" into a tibble
library(tidyverse)
res <- as_tibble(resData)
## save for viz
saveRDS(res, file = 'data/rvecRslurmMapRes.rds')
## cleanup careful this deletes output also
## cleanup_files(sjob) 
```

```{r ch030-rvecRslurmMapFig, eval=FALSE, out.width= '75%', fig.cap='The continous marginal pairs exact reproduce the specified correlation whereas the discrete pairs (green nad blue) show a downward bias. Future work with employ a discrete adjusted procedures using *rescaling* on the traditional correlation coefficient.'}
res <- readRDS( 'data/rvecRslurmMapRes.rds' )
res
res %>%
    ggplot( aes( x = rho, y = rhoHat, color = pair ) )  +
    geom_point() +
    geom_abline( slope = 1, intercept = 0 )
```

>>>>

This section describes a low-dimensional (2D) random vector simulation workflow via the `bigsimr` `R` package.
`bigsimr` provides an interface to the native code written in Julia, registered as the `Bigsimr` Julia package PROVIDE URL.
In addition to the native Julia `Bigsimr` package and `R` interface `bigsimr`, we also provide a python interface `bigsimr` (PROVIDE URL) that interfaces with the Julia `Bigsimr` package. 
The Julia implementation provides a high-performance implementation of our proposed random vector generation algorithm and associated functions (see Section \@ref(algorithms)).
When designing `bigsimr`, we aimed to conveniently provide parallelized computation, through multi-core and GPU acceleration, while allowing advanced users to customize.

The subsections below describe the basic use of `bigsimr`, by stepping through an example workflow using the data set `airquality` that contains daily air quality measurements in New York, May to September 1973 [@Chambers1983].
This workflow proceeds from computing environment setup, data wrangling, estimation, simulation configuration, random vector generation, and, finally, result visualization.
The code chunk below prepares the computing environment:

@@

## Bivariate example description

We illustrate the use of `bigsimr` motivated by the New York air quality data set (`airquality`) included in the R `datasets` package.
First, we load the `bigsimr` library and a few other convenient data science packages, including the syntactically-elegant `tidyverse` suite of `R` packages.

@@

## Specifying marginal distributions

Based on the estimated densities in Figure \@ref(fig:ch030-aq-joint-dist), we assume `Temp` is normally distributed and `Ozone` is log-normally distributed, as the latter values are positive and skewed. We use the well-known, unbiased estimators for the normal distribution's parameters and maximum likelihood estimators for the log-normal parameters:

@@

## Specifying correlation

As mentioned, the user must decide how to describe correlation, based on the particulars of the problem. For non-normal data and for improved simulation accuracy/scalability in our scheme, we advocate the use of Spearman's $\rho$ correlation matrix $R_S$ and Kendall's $\tau$ correlation matrix $R_K$. We also support Pearson correlation coefficient matching, while cautioning the user to check the performance for their parametric multivariate model (see [Monte Carlo evaluations](#simulations) below for evaluation strategies and guidance). To aid in correlation specification, and estimation in general, we provide a high-performance function `cor_fast` which estimates Spearman, Kendall, or Pearson correlation of large data sets by computing pairwise correlations in parallel. (Anyone who has tried estimating Kendall's $\tau$ using `stats::cor` can attest that the routine does not scale to even moderate dimensions). These estimation methods are classical approaches, not designed for high-dimensional correlation estimation (see the [Conclusion and Discussion]({#discussion) sections for more on this).

@@

## Checking target correlation matrix admissibility

First we use `cor_bounds` to ensure that the pairwise target correlation values are valid prior to mapping step which constructs $R_X$ for the MVN input into NORTA (see Section \@ref(algorithms)). `cor_bounds` estimates the pairwise lower and upper correlation bounds using the Generate, Sort, and Correlate algorithm of @DH2011.

@@

## Simulating random vectors
Finally, we arrive at the main function of `bigsimr`, `rvec`. We now simulate $B=10,000$ random vectors from the assumed joint distribution of Ozone levels and Temp.

@@
